Evaluating Deep Learning Methods for Tokenization of Space-less texts in Old French and Latin

### 2019-06-07 article/article.tex
Updated the article with style correction, moved annex into Evaluation,, etc.

\title{Evaluating Deep Learning Methods for Tokenization of Space-less texts in Old French}
\author[1]{Thibault Clérice}
\affil[1]{École nationale des Chartes, France}
\affil[2]{Université Lyon 3, France}

@@

\abstract{Tokenization of modern and old Western European languages seems to be fairly simple as it stands on the presence mostly of markers such as spaces and punctuation. Although, when dealing with old sources like manuscript written in \textit{scripta continua} or later manuscripts, (1) such markers are mostly absent, (2) spelling variation and rich morphology makes dictionary based approaches difficult. We show that applying convolutional encoding to characters followed by linear categorization to word-boundary or in-word-sequence can be used to tokenize such inputs. Additionally, we release a software with a rather simple interface for tokenizing one's corpus.}

@@

% To Read : Stutzmann article.

Tokenization of space-less strings is a task that is specifically difficult for computer when compared to "whathumancando". \textit{Scripta continua} is a writing phenomenon where words are be separated by spaces and has disappeared around the 8th century (see \citet{zanna1998lecture}). Never the less, spacing can be somewhat erratic in later centuries writings as \citet{stutzmann} has shown (\textit{cf.} Figure \ref{fig:4lines}, a document from the 13th century, displays this kind of situation). In the context of text mining of HTR or OCR output, lemmatization and tokenization of medieval western languages can be a pre-processing step for further research to sustain analyses such as authorship attribution or simply allow full-text search.

@@

We must stress in this study that the difficulty that we face is different for \textit{scripta continua} than the ones researchers face languages such as Chinese for which an already impressive amount of work has been done as it. Indeed, Chinese word segmentation has lately been driven by deep learning methods, specifically ones based on \textit{sequence to sequence translations}: \citet{chen2015long} defines a process based on LSTM model, while \citet{yu2019learning} uses BiDirectional GRU and CRF. Actually, meanwhile redacting this article and producing the code-base, \citet{huang2019realistic} took the same approach of encoding to linear classification to both word boundary (WB) and word content (WC) for Chinese word segmentation.

Indeed, while Chinese's issue seems to lie in the decomposition of relatively fix characters, Old French or medieval latin present heavy variation of spelling. In \citet{camps_pandora}, Camps notes, in the same corpus, the existence of not less than 29 spelling of the word \textit{cheval} (horse in Old and Modern French) whose apparition counts span from 3907 to 1\footnote{These are \textit{cheval}, \textit{chevaus}, \textit{cheual}, \textit{ceval}, \textit{chevals}, \textit{cevaus}, \textit{chival}, \textit{ceual}, \textit{cheuaus}, \textit{cevals}, \textit{chaval}, \textit{chivaus}, \textit{chiual}, \textit{chevas}, \textit{cheuals}, \textit{chiuaus}, \textit{ceuaus}, \textit{chevaul}, \textit{chiuau}, \textit{chivals}, \textit{chevau}, \textit{kevaus}, \textit{chavaus}, \textit{cheuas}, \textit{keval}, \textit{cheua}, \textit{cheuau}, \textit{cheva}, \textit{chiuals}}. This  makes a dictionary approach rather difficult as it would rely on a high number of different spelling and makes the computation highly complex.

@@

For evaluation purposes, and to reduce the number of input classes, we propose two options for data transcoding: a lower-case normalization and a "reduction to the ASCII character set" feature (fr. \ref{fig:normalization}). On this point, a lot of issues were found with transliteration of medieval paelographic characters that were part of the original datasets, as they are badly interpreted by the \texttt{unidecode} python package. Indeed, \texttt{unidecode} will simply remove characters it does not understand. I built a secondary package named \texttt{mufidecode} (\citet{thibault_clerice_2019_3237731}) which precedes unidecode equivalency tables when the data is known of the Medieval Unicode Font Initiative (MUFI, \citet{mufi}).

@@

Aside from normalizations of the input and output, three different structure of models were tested. Every model is composed by one encoder described below and one Linear Classifier which classifies into 5 classes : Start of Sentence (= SOS), End of Sentence (= EOS), Padding (= PAD), Masked Token (= Word Content), Space (= Word Boundary). For final scores, SOS, EOS and PAD were ignored.

@@

Datasets are transcription from manuscripts with unresolved abbreviation coming from different projects. The \textbf{Old French} is based on \citet{8269990}, \citet{pinche:hal-01628533}, \citet{jean_baptiste_camps_2019_2630574}, \citet{bfmmss}, and \citet{tnah_transcription}. It contains

@@

The input was generated by grouping at least 2 words and a maximum of 8 words together per sample. On a probability of 0.2, noise character could be added (noise character was set to DOT ('.')) and some words were kept randomly from a sample to another on a probability of 0.3 and a maximum number of word kept of 1. If a minimum size of 7 characters was not met in the input sample, another word would be added to the chain. A maximum input size of 100 was kept. The results corpora should be varied in sizes as shown by \ref{fig:word_sizes}. The corpora is composed by 193 different characters when not normalized, in which some MUFI characters appears few hundred times \ref{tab:mufi_examples}.

@@

The training parameters was 0.00005 in learning rate for each CNN model and 0.001 for the LSTM one, and 64 in batch sizes. Training reached a plateau fairly quickly for each model (\textit{cf.} \ref{fig:loss}). Each model except LSTM reached a really low loss and a high accuracy on the test set (\textit{cf.} \ref{tab:scores}). To compare the results, we used the \texttt{wordsegment} package \citet{WordSegment} as a baseline.

@@

\begin{table}[!ht]
\centering
\begin{tabular}{lllll}
\hline
Model    & Accuracy & Precision & Recall & FScore \\ \hline
Baseline & 0.989    & 0.986     & 0.984  & 0.985 \\
CNN      & 0.991    & 0.985     & 0.990  & 0.987  \\
CNN L    & 0.991    & 0.979     & 0.990  & 0.985  \\
CNN P    & \textbf{0.993}    & \textbf{0.990}& \textbf{0.991}  & \textbf{0.990}  \\
CNN N    & 0.991    & 0.987     & 0.988  & 0.988  \\
CNN L N  & 0.992    & 0.988     & 0.989  & 0.988  \\
LSTM     & 0.741    & 0.184     & 0.500  & 0.269  \\ \hline
\end{tabular}
\caption{Scores over the test dataset. N = normalized, L = Lower, P = no position embedding.}
\label{tab:scores}
\end{table}

@@

While all model using CNN shows improvement over the baseline, the model definitely does not outperform it by a huge margin (\textless 0.02 FScore). And for a reason : the baseline already performs nearly perfectly on the test corpus. The dictionary attack using n-grams did actually perform well. As a result, we wanted to compare how both models would perform on a secondary test corpus composed by texts that were not used in training: indeed, the training, dev and test corpus share the same texts while not sharing the same inputs. As a result, we created a new corpus with 4 texts and 742 samples : the diplomatic edition of the \textit{Graal} (\citet{graal}), a \textit{Passion} and a \textit{Vie de Saint Leger}  (\citet{old_french_corpus}), a \textit{Vie de Saint Thibaut} (\citet{theobaldus}). No noise characters and no random keeping of words were applied.

@@

The following inputs has been tagged with the CNN P model. Batch are constructed around the regular expression \texttt{\\W} with package \texttt{regex}. This explains why inputs such as \texttt{".i."} are automatically tagged as \texttt{" . i . "} by the tool. The input was stripped of its spaces before tagging, we only show the ground truth by commodity.

@@

@@

We believe that, aside from a graphical challenge, word segmentation in OCR from manuscripts can actually be treated from a text point of view and as a NLP task. Word segmentation for some text can be even difficult for humanist, and as such, we believe that post-processing of OCR through tools like Boudams can be a better way to achieve data-mining of the dataset.

We were surprised by the negligible effects of the different normalization methods (lower-casing; ASCII reduction; both). The presence of certain MUFI characters might provide enough information about segmentation and be in enough numbers for them not to impact the network weights.

While the baseline surprised us by performing this well on the test corpus, it definitely performed less well than the CNN on a completely unknown corpus: in this context, the proposed model actually shows its ability to carry over unknown corpora in a better way than classical ngram approaches. In light of the high accuracy of the CNN model, we believe the model should perform the same way independently from the language in Medieval Western Europe, and results in annexes on Latin confirms this for at least 2 other corpora.

@@

Achieving 0.99 accuracy on word segmentation with a corpus as large as 25,000 test samples seems to be the first step for a more important data mining of OCRed manuscript. In aftermath, we wonder if the importance of normalization and lowering should be higher depending on the size of the corpora and its content.

@@

\section{Annex 2 : Scores on Latin Prose and Poetic Corpora}

The Latin data is much more noisy than the Old French, as it was less curated than the digital edition we found for Old French. They are part of the Perseus corpus \citet{perseus} and were cut into passages in the context of my thesis. The training, evaluation and test corpora are built upon prose work from Cicero and Suetonius. The unknown corpus is built upon \textit{Epigrammata} from Martial, from book 1 to book 2, as it should be fairly different in word order, vocabulary, etc. Both corpus were generated without noise and word keeping, with a maximum sample size of 150 characters.

\section{Annex 3: Scores on Medieval Latin corpora}

\begin{itemize}
\item Number of training examples: 36814
\item Number of evaluation examples: 4098
\item Number of testing examples: 5612
\item Number of classes in testing examples: 137,465 WC; 34,053 WB
\item Number of classes in unknown examples: 472,655 WC; 113,004 WB
\end{itemize}

\section{Annex 4: Scores on Latin epigraphic corpora}

The epigraphic Latin corpora is based on the Epigraphic Database Heidelberg open data \citet{formulae} for its training, evaluation and test sets (HD000001-HD010000 and HD010001-HD020000 from \citet{edh}) while the corpus of unknown is drawn from an automatic conversion of the Pompei Inscriptions (\citet{pompei}). We also decided to evaluate both the baseline and the model on uppercased data, as it would normally be the state the text would be found in. Each of the corpora presents a high number of unresolved abbreviations (\textit{ie.} one letter words). Both corpus were generated without noise and word keeping, with a maximum sample size of 150 characters. The data presents some polytonic Greek characters, some sample being only in Greek.

While CNN is influenced by
\textbf{Statistics:}

>>>>

\title{Evaluating Deep Learning Methods for Tokenization of Space-less texts in Old French and Latin}
\author[1]{Thibault Clérice}
\affil[1]{École nationale des Chartes, France}
\affil[2]{Université Lyon 3, France}

@@

\abstract{Tokenization of modern and old Western European languages seems to be fairly simple, as it stands on the presence mostly of markers such as spaces and punctuation. However, when dealing with old sources like manuscripts written in \textit{scripta continua}, antiquity epigraphy or Middle Age manuscripts, (1) such markers are mostly absent, (2) spelling variation and rich morphology make dictionary based approaches difficult. Applying convolutional encoding to characters followed by linear categorization to word-boundary or in-word-sequence is shown to be effective at tokenizing such inputs. Additionally, the software is released with a simple interface for tokenizing a corpus or generating a training set.}

@@

Tokenization of space-less strings is a task that is specifically difficult for computers as compared to "whathumanscando". \textit{Scripta continua} is a writing phenomenon in which words are separated by spaces that disappeared around the 8th century (see \citet{zanna1998lecture}). Nevertheless, spacing can be somewhat erratic in later centuries writings as \citet{stutzmann} explains (\textit{cf.} Figure \ref{fig:4lines}) and it becomes an issue for OCR, as continuous bag of word is only interesting when those are not glued together. In the context of text mining of HTR or OCR output, lemmatization and tokenization of medieval western languages can be a pre-processing step for further research to sustain analyses such as authorship attribution or simply allow full-text search.

@@

It must be stressed in this study that the difficulty inherent to segmentation is different for \textit{scripta continua} than the one for languages such as Chinese for which an already impressive amount of work has been. Chinese word segmentation has lately been driven by deep learning methods, specifically ones based on \textit{sequence to sequence translations}: \citet{chen2015long} defines a process based on LSTM model, while \citet{yu2019learning} uses BiDirectional GRU and CRF. \footnote{\citet{huang2019realistic} actually gave us the denomination used here: word boundary (WB) and word content (WC).}

Indeed, while the issue with Chinese seems to lie in the decomposition of relatively fixed characters, Old French or Medieval Latin present heavy variation of spelling. In \citet{camps_pandora}, Camps notes, in the same corpus, the existence of not less than 29 spellings of the word "\textit{cheval}" (horse in Old and Modern French) whose apparition counts span from 3907 to 1\footnote{These are \textit{cheval}, \textit{chevaus}, \textit{cheual}, \textit{ceval}, \textit{chevals}, \textit{cevaus}, \textit{chival}, \textit{ceual}, \textit{cheuaus}, \textit{cevals}, \textit{chaval}, \textit{chivaus}, \textit{chiual}, \textit{chevas}, \textit{cheuals}, \textit{chiuaus}, \textit{ceuaus}, \textit{chevaul}, \textit{chiuau}, \textit{chivals}, \textit{chevau}, \textit{kevaus}, \textit{chavaus}, \textit{cheuas}, \textit{keval}, \textit{cheua}, \textit{cheuau}, \textit{cheva}, \textit{chiuals}}. This  makes a dictionary-based approach rather difficult as it would rely on a high number of different spellings, making the computation highly complex.

@@

For evaluation purposes, and to reduce the number of input classes, two options for data transcoding were used: a lower-case normalization and a "reduction to the ASCII character set" feature (fr. \ref{fig:normalization}). On this point, a lot of issues were encountered with transliteration of medieval paelographic characters that were part of the original datasets, as they are poorly interpreted by the \texttt{unidecode} python package. Indeed, \texttt{unidecode} will simply remove characters it does not understand. A derivative package named \texttt{mufidecode} was built for this reason(\citet{thibault_clerice_2019_3237731}): it takes precedent over \texttt{unidecode} equivalency tables when the data is known of the Medieval Unicode Font Initiative (MUFI, \citet{mufi}).

@@

Aside from normalizations of the input and output, three different model structures were tested. Every model is composed of one encoder, as described below, and one Linear Classifier which classifies into 5 classes : Start of Sentence (= SOS), End of Sentence (= EOS), Padding (= PAD), Masked Token (= Word Content), Space (= Word Boundary). For final scores, SOS, EOS and PAD were ignored.

@@

The dataset is composed of transcriptions (from different projects) of manuscripts with unresolved abbreviation. The \textbf{Old French} is based on \citet{8269990}, \citet{pinche:hal-01628533}, \citet{jean_baptiste_camps_2019_2630574}, \citet{bfmmss}, and \citet{tnah_transcription}. It contains

@@

Examples were generated automatically. They are between 2 and 8 words-length. In order to recreate the condition of OCR noise, dots were added randomly (0.2) between words. In order to augment the dataset, words are randomly (0.1) passed over the next example\footnote{This data augmentation was limited to one word per sample}. If a minimum size of 7 characters was not met in the input sample, another word would be added to the chain, independently of the maximum number of words. The example however should not go beyond 100 characters. The results corpora should be varied in sizes as shown by Figure \ref{fig:word_sizes}. The corpora contains 193 different characters when not normalized, in which some MUFI characters appears few hundred times (\textit{cf.} Table \ref{tab:mufi_examples}).

@@

The training parameters were 0.00005 in learning rate for each CNN model, 0.001 for the LSTM model, and batch sizes of 64. Training reached a plateau fairly quickly for each model (\textit{cf.} \ref{fig:loss}). Each model except LSTM reached a really low loss and a high accuracy on the test set (\textit{cf.} \ref{tab:scores}). To compare the results, the \texttt{wordsegment} package \citet{WordSegment} was used as a baseline.

@@

\begin{table}[!ht]
\centering
\begin{tabular}{lllllll}
\hline
Model    & Accuracy & Precision & Recall & FScore & WB FN & WB FP \\ \hline
Baseline & 0.989    & 0.986     & 0.984  & 0.985 & 4031 & 3229 \\
CNN      & 0.991    & 0.985     & 0.990  & 0.987 & 2137 & 3860 \\
CNN L    & 0.991    & 0.979     & 0.990  & 0.985 & \textbf{2117} & 3750 \\
CNN P    & \textbf{0.993}    & \textbf{0.990}& \textbf{0.991}  & \textbf{0.990} & 2432 & \textbf{2114} \\
CNN N    & 0.991    & 0.987     & 0.988  & 0.988 & 2756 & 3312 \\
CNN L N  & 0.992    & 0.988     & 0.989  & 0.988 & 2500 & 3567 \\
LSTM     & 0.741    & 0.184     & 0.500  & 0.269 & 169094 & 0 \\ \hline
\end{tabular}
\caption{Scores over the test dataset. \\\hspace{\textwidth}For models: N = normalized, L = Lower, P = no position embedding. \\\hspace{\textwidth}In headers, FN = False Negative, FP = False Positive}
\label{tab:scores}
\end{table}

@@

While all models using CNN show improvement over the baseline, the models definitely do not significantly outperform (\textless 0.02 FScore). There is a reason for this: the baseline already performs nearly perfectly on the test corpus. The dictionary attack using n-grams did actually perform well. Therefore, an additional evaluation method was constructed. The baseline and the best achieving deep-learning model (CNN P) were evaluated on a secondary test corpus composed of texts that were not used in training: indeed, the training, dev and test corpus share the same texts while not sharing the same source corpora and as such the same writing, the same vocabulary. This new corpus is composed by 4 texts and counts 742 examples : the diplomatic edition of the \textit{Graal} (\citet{graal}), a \textit{Passion} and a \textit{Vie de Saint Leger}  (\citet{old_french_corpus}), a \textit{Vie de Saint Thibaut} (\citet{theobaldus}). Neither noise characters nor random keeping of words were applied.

@@

The following inputs have been tagged with the CNN P model. Batches are constructed around the regular expression \texttt{\\W} with package \texttt{regex}. This explains why inputs such as \texttt{".i."} are automatically tagged as \texttt{" . i . "} by the tool. The input was stripped of its spaces before tagging, only the ground truth is shown for readability.

@@

\subsection{Evaluation on Latin data}

For the following evaluations, the same process was deployed: CNN without Position was evaluated against the baseline on both a test set composed by the same texts that the training text, and an unknown corpora composed by unseen texts.

\subsubsection{Latin Prose and Poetic Corpora}

The Latin data is much more noisy than the Old French, as it was less curated than the digital edition provided for Old French. They are part of the Perseus corpus \citet{perseus} and were cut into passages in the context of my thesis. The training, evaluation and test corpora are built upon prose works from Cicero and Suetonius. The unknown corpus is built upon \textit{Epigrammata} from Martial, from book 1 to book 2, as it should be fairly different in word order, vocabulary, etc. Both corpus were generated without noise and word keeping, with a maximum sample size of 150 characters.

\subsubsection{Medieval Latin corpora}

\begin{itemize}
\item Number of training examples: 36814
\item Number of evaluation examples: 4098
\item Number of testing examples: 5612
\item Number of classes in testing examples: 137,465 WC; 34,053 WB
\item Number of classes in unknown examples: 472,655 WC; 113,004 WB
\end{itemize}p

\subsubsection{Latin epigraphic corpora}

The epigraphic Latin corpora is based on the Epigraphic Database Heidelberg open data \citet{formulae} for its training, evaluation and test sets (HD000001-HD010000 and HD010001-HD020000 from \citet{edh}) while the corpus of unknown is drawn from an automatic conversion of the Pompei Inscriptions (\citet{pompei}). Both the baseline and the model were evaluated on uppercase data, as it would normally be the state the text would be found in. Each of the corpora presents a high number of unresolved abbreviations (\textit{ie.} one letter words). Both corpus were generated without noise and word keeping, with a maximum sample size of 150 characters. The data presents some polytonic Greek characters, some sample being only in Greek.

@@

Aside from a graphical challenge, word segmentation in OCR from manuscripts can actually be treated as a NLP task. Word segmentation for some text can be even difficult for humanist, as shown by the manuscript sample, and as such, it seems that the post-processing of OCR through tools like this one can be a better way to achieve data-mining of raw datasets.

The negligible effects of the different normalization methods (lower-casing; ASCII reduction; both) were surprising. The presence of certain MUFI characters might provide enough information about segmentation and be of sufficient quantity for them not to impact the network weights.

While the baseline performed unexpectedly well on the test corpus, the CNN model definitely performed better on a completely unknown corpus. In this context, the proposed model actually shows its ability to carry over unknown corpora in a better way than classical n-gram approaches. In light of the high accuracy of the CNN model, the model should perform the same way independently of the language in Medieval Western Europe,.

@@

Achieving 0.99 accuracy on word segmentation with a corpus as large as 25,000 test samples seems to be the first step for a more thorough data mining of OCRed manuscript. Given the results, studying the importance of normalization and lowering should probably be a further step, as it might be of high influence in smaller corpora.

@@
